---
title: "Interpretation of Linear Regression Models with Abundance-Based Constraints"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpretation of Linear Regression Models with Abundance-Based Constraints}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lmabc)
```

```{r generate_data, include=FALSE}
n <- 200

groups_1 <- c("Asian", "Hisp", "NHB", "NHW")
pi_1 <- c(0.10, 0.15, 0.20, 0.55)
K_1 <- length(groups_1)
k_1 <- factor(sample(groups_1,
                    size = n, replace = TRUE,
                    prob = pi_1),
             levels = c("NHW", "Asian", "Hisp", "NHB"))

x1 <- rnorm(n = n, mean = 5,  sd = 2)

Ey <- 1 + 2*x1 + 3*I(k_1==groups_1[1]) -.5*I(k_1==groups_1[2]) + .5*I(k_1==groups_1[2])*x1
y <- Ey + rnorm(n=n)

df <- data.frame(y, age = x1, race = k_1)
```

## Motivation: inequitable interpretation
The inequities in traditional baseline-encoded linear regression are easy to see with model output in R. Let's look at an example with some simulated data:

```{r}
summary(lm(y ~ age + race + age:race, df))
```

The output does not explicitly note that the continuous effect (`age`) is specific to the NHW race group and not a global effect for the population; similarly, the `age:Hisp` and `age:NHB` effects do not actually correspond to the race-specific slopes, but rather the _difference_ between the NHW-specific slope and the race-specific slopes for Hispanic and NH Black individuals, respectively.

Furthermore, regularization methods like variable selection would unfairly shrink coefficients toward the NHW-specific effect. The (racial) asymmetry is clear, and model-based inferences may differ depending on the choice of baseline.

Together, we see that the default "baseline" method of linear regression creates inequities in interpretation and analysis. We believe it is important to deploy a more fair strategy, such as linear regression with abundance-based constraints.

## Our solution: abundance-based constraints
In the corresponding paper to this R package, we explain in great depth the mathematical properties of our `lmABC` operation. This vignette will briefly explain model interpretation.

Consider a simple model using the data from the previous section. Now, let's use `lmABC`.

```{r}
fit_no_int <- lmabc(y ~ age + race, df)
summary(fit_no_int)
```

In this new model output, `age` corresponds to the global effect, and the other slopes modify the global effect. For instance, being in the NHW category shifts the intercept by `r round(coef(fit_no_int)[["raceNHW"]], 3)` units in $y$ compared to the population as a whole.

We can add interaction terms.

```{r}
fit_int <- lmabc(y ~ age + race + age:race, df)
summary(fit_int)
```

Here, the interaction terms modify the coefficient of `age`, which is now the global effect of age on $y$.

Crucially, the model itself is the same as the traditional baseline-encoded linear regression model. Thus, all properties of the model—residuals, R-squared, information for diagnostics—remain the same. `lmABC` just computes a different, fairer expression of the model than `lm` provides.

The `lmABC` package includes implementations for many common methods: `summary`, `print`, `plot`, `predict`, `logLik`, `vcov`, and more. And we don't just have the abundance-based constraints version of `lm`—check out `glmABC`, too!
