---
title: "Limitations of lmABC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Limitations of lmABC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lmabc)
```

As a method still under development, our implementation of `lmabc` is not exhaustive. This vignette will serve as a running list of limitations of `lmabc`. Email the package maintainer with questions or suggestions.

## `lmabc` throws an error for missing factor levels
Consider an interaction between two categorical variables, `k1` and `k2`, with levels `a`, `b`, and `c` and `x`, `y`, and `z`, respectively. If any pair has no observations—for instance, if there are no observations with `k1 = a`, and `k2 = z`—then `lmabc` will throw an error.

The standard, baseline-encoded `lm` ignores this level and returns an `NA` for its coefficient. However, we do not believe this is a good solution.

## Data must be passed as a data.frame
The standard implementation of `lm` can handle data passed as vectors, thus rendering the `data` argument optional. `lmabc` requires the columns referenced in both sides of the formula to be passed to `lmabc` together in a single data frame.

## Missing implementations of generic methods
We implemented many of the common generic methods associated with linear regression models, but many remain, especially in non-base packages. `lmabc` objects only have class "lmabc," so methods only implemented for class "lm" will fail unless the generic method has a default.

We will implement additional generic methods throughout the development process.

## Outcome types
For `lmabc`, the outcome variable must be continuous. Further, there may only be one outcome variable.
