---
title: "Limitations of lmabc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Limitations of lmabc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lmabc)
```

The statistical methods, theory, and algorithms for regression with Abundance-Based Constraints (ABCs) are under active development. As such, the current implementation of `lmabc` is not exhaustive. This vignette will serve as a running list of limitations of `lmabc`. Please email the package maintainer with questions or suggestions.

## `lmabc` throws an error for missing factor levels
Suppose we have two categorical variables, `k1` and `k2`, with levels `a`, `b`, and `c` and `uu`, `vv`, respectively. If either `k1` or `k2` has zero observations in at least one level, then a model of the form `y ~ k1 + k2 + ...` will throw an error. Similarly, if an interaction is included, `y ~ k1 + k2 + k1:k2 + ...`, but a joint category (e.g., `k1 = a` and `k2 = uu`) has zero observations, then `lmabc` will throw an error.

This behavior is different from `lm`, which instead removes each unobserved level from the model and returns an `NA` for its coefficient. 

## Data must be passed as a data.frame
The standard implementation of `lm` can handle data passed as vectors, thus rendering the `data` argument optional. `lmabc` requires all variables in the formula statement to be passed to `lmabc` as a single data frame.

## Missing implementations of generic methods
We implemented many of the common generic methods associated with linear regression models. However, our implementation is not exhaustive, especially for non-base packages. `lmabc` objects only have class "lmabc", so methods only implemented for class "lm" will fail unless the generic method has a default.

We will implement additional generic methods throughout the development process.

